---
title: "overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Moran Eigenvector Spatial Filtering

## About

The purpose of this package is to implement Moran Eigenvector Spatial Fitlering (MESF),
a technique used for reducing spatial autocorrelation in generalized linear models (GLMs). 
MESF searches eigenvectors of a transformed spatial weights matrix (a doubly 
centered matrix). One-by-one, eigenvectors are added to the right-hand side of the 
GLM and Moran's I statistic (a measure of spatial autocorrelation) is calculated 
for the residuals of the newly fitted model. The eigenvector which produces the 
lowest Moran's I is then permanently added to the model and the search process 
repeated to choose additional eigenvectors to add to the model. This process continues 
until the p-value for Moran's I exceeds the specified stopping threshold alpha.


## Example Regression Problem

```{r setup, message=FALSE}

# packages needed for demo
library(sp)
library(spdep)
library(MESFfast)
```

We will use data from Columbus, OH to demonstrate how to build a regression model
with MESF. In this model, we'll try to predict crime rates based on a neighborhood's
average home value and household income level.

The data set columbus_sp is provided in this package. It is a SpatialPointsDataFrame 
object, with latitude and longitude columns. We'll load in the data set, then
create a distance weighting matrix with weights based on the inverse distance
between two points.

```{r}

# sample spatial data set with crime stats by neighborhood for Columbus, OH
data(columbus_sp)
```

We first show what a model would look like without applying the MESF method. This
model aims to predict crime rates using some attributes of the neighborhood: the
average home value, and the distance of the neighborhood to the central business 
district (downtown area). 

```{r}

# fit a standard linear model
basic_model <- glm(crime ~ home_value + income, data = columbus_sp)
summary(basic_model)
```
Both covariates are predictive of crime rates. But wait! Before we finalize our
model, let's check for some potential issues. One concern with spatial data
is spatial autocorrelation in the residuals, which can lead to biased parameter
estimates and underestimation of standard errors.

We can use a function from this package to assess spatial autocorrelation in the
residuals. We calculate Moran's I statistic and the associated p-value. This test
checks our data against the null hypothesis that the residuals from our regression
model are randomly distributed, with no spatial clustering or dispersion.

```{r}

# calculate Moran's I
moranfast(basic_model$residuals, lat=columbus_sp$lat, lon=columbus_sp$lon)
```
Possible values for Moran's I range from -1 (spatial dispersion) to 0 (random 
distribution) to 1 (spatial clustering). We observe a Moran's I of around 
$I=0.09$, which indicates a small degree of spatial clustering, meaning
neighborhoods near one another are more likely to have residuals which are 
similar to one another. There is extremely strong evidence 
($p \approx 8\times10^{-7}$) that the observed Moran's I would not be generated
through random variation in residuals.

We can address this spatial autocorrelation in the residuals by applying the
MESF method. We will test out the largest 10 eigenvectors of the spatial weights
matrix. We do this in verbose mode so we can get better insight into what the
function is doing as it runs.

```{r}

# identifies neighbors of points
# looks for neighbors between 0.01 and 75 km away
neighbors <- dnearneigh(columbus_sp, d1 = 0.0001, d2 = 75) 

# create inverse distance based spatial weights for list of neighbors
idw_weights <- nb2listwdist(neighbors, columbus_sp, type = "idw", zero.policy=TRUE, style="W")

# find moran eigenvector spatial filters to use
find_mesf <- mesf(crime ~ home_value + income,
                   data = columbus_sp,
                   family = gaussian,
                   listw = idw_weights, verbose = T, alpha = .05,
                   n_eigs = 10, lat = columbus_sp$lat, lon = columbus_sp$lon,
                   parallel = FALSE) # we set parallel to false only for the purposes of the vignette
```

If we look at the output from the function, we can preview the eigenvectors which
the fitting process has recommended to include in our final model.

```{r}

# preview the eigenvectors chosen
head(find_mesf$vectors)
```
It looks like there are 3 eigenvectors we'd like to include in our final model.
So we can re-fit our regression, adding in these eigenvectors to the right-hand
side. 

```{r}

# refit regression
final_model <- glm(crime ~ home_value + income 
                   + find_mesf$vectors[,1] + find_mesf$vectors[,2] + find_mesf$vectors[,3],
                   data = columbus_sp)
summary(final_model)
```
Now we can check for the spatial autocorrelation in the residuals of our final model.

```{r}

# calculate Moran's I on final model
moranfast(final_model$residuals, lat=columbus_sp$lat, lon=columbus_sp$lon)
```
We have reduced the Moran's I and note that the p-value exceeds the $\alpha$ threshold
we set, which was 0.05.

